---
title: "Land_class"
author: "Enrico Lucci"
date: "2023-04-20"
output:
  pdf_document: default
  html_document:
    df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

## Classifying physical landscape
References for landforms classification: 
- Meybeck et al (2001)
- https://esdac.jrc.ec.europa.eu/projects/landform-classification

Basing on landscape classification  Meybeck et al (2001), I classified landforms according 
to the elevation and the Topographic Roughdenss Indes (TRI).



#### Classifying landscape according to elevation values
In this step I classified the physical landscape according to elevation values, 
using the following classes:

1) 0-100
2) 100-200
3) 200-500
4) 500-1000
5) 1000-1500
6) 1500-inf


```{r libraries}

library(raster)
library(terra)
library(spatialEco)
library(RColorBrewer)
```


#### Study area

```{r study area, echo=TRUE, warning=FALSE}


sa <- vect("~/GIS_ModLandWGS84/shapefile/studyarea.shp")
plot(sa)

```




```{r elevation raster analysis, echo=TRUE, message=FALSE, warning=FALSE}


elev <- rast("~/GIS_ModLandWGS84/raster/Abr_Mol_Pug.tif")
# elev <- aggregate(elev, fact = 3) funzione che permnette di ridurre la definizione del raster 
# ma perde di efficacia l'analisi TRI, non restituendo un raster utile per definire l'asperità del terreno  
plot(elev, main = "elevation (m asl)")

# Generate a reclass matrix (begin, end, value ...)
m = c(-Inf, 100, 1, 100, 200, 2, 200, 500, 3, 500, 1000, 4, 1000, 1500, 5, 
      1500, Inf, 6)
rclmat = matrix(m, ncol=3, byrow=TRUE)

# Reclassify the raster layer
reclassified_elev = classify(elev, rclmat)
reclassified_elev[reclassified_elev < 0] <- NA #exclude negative values

sum(reclassified_elev)
hist(reclassified_elev, main = "Elevation classes")
plot(reclassified_elev, main = "elevation classes (m asl)", col=topo.colors(6), legend = FALSE)
  legend(13,40.9, title="classes",
   c("0-100","100-200","200-500","500-1000","1000-1500",">1500"), 
   fill=topo.colors(6), cex=0.8)

  
  #writeRaster(reclassified_elev, '~/R analisi/ModLand/data/raster/rec_elev.tiff', overwrite=TRUE)
  
  
```


#### Classifying landscape according to TRI values
 I used the TRI classification of the SpatialEco package


Recommended ranges for classifying Topographic Ruggedness Index:

0-80 - level terrain surface.

81-116 - nearly level surface.

117-161 - slightly rugged surface.

162-239 - intermediately rugged surface.

240-497 - moderately rugged surface.

498-958 - highly rugged surface.

over 959 - extremely rugged surface.




```{r TRI, echo=TRUE, message=FALSE, warning=FALSE}


elev_rid <- aggregate(elev, fact = 3)

TRI <- tri(elev_rid, exact = TRUE, s=5) #max s = 5


sum(TRI)
density(sum(TRI))
plot(TRI, main = "TRI by spatialEco")

# Reclassify the TRI raster layer

TRI_classes = c(-Inf, 80, 1, 80, 116, 2, 116, 161, 3, 161, 239, 4, 239, 497, 5,497,Inf,6)

TRIclmat = matrix(TRI_classes, ncol=3, byrow=TRUE)

reclassified_TRI <- classify(TRI, TRIclmat)
sum(reclassified_TRI)
hist(reclassified_TRI, main = "TRI classes")

plot(reclassified_TRI, main = "TRI classes", col=topo.colors(6), legend = FALSE)
  legend(13,40.9, title="classes",
   c("level terrain surface","nearly level surface","slightly rugged surface","intermediately rugged surface","moderately rugged surface", "highly rugged surface"), 
   fill=topo.colors(6), cex=0.8)
  
  #writeRaster(reclassified_TRI, '~/R analisi/ModLand/data/raster/rec_TRI.tiff', overwrite=TRUE)

```

#### Polygonizing raster layers of elev and TRI.
To transform the raster layers (i.e. elevation and the TRI) I use the 
function as.polygon of the package "Terra", but I need firstly to change 
"RasterLayer" objects in "Formal class SpatRaster" by the package R

```{r from raster to polygon, message=FALSE, warning=FALSE}

#transforming "RasterLayer" objects in "Formal class SpatRaster" by the package R

reclassified_elev <- rast("~/R analisi/ModLand/data/raster/rec_elev.tiff")
reclassified_TRI <- rast("~/R analisi/ModLand/data/raster/rec_TRI.tiff")

ec_vec <- as.polygons(reclassified_elev)
TRI_vec <- as.polygons(reclassified_TRI)

plot(ec_vec)
plot(TRI_vec)

elev_TRI <- union(ec_vec, TRI_vec)
plot(elev_TRI)

#writeVector(elev_TRI, '~/R analisi/ModLand/data/elev_TRI.shp', overwrite=TRUE)

```

#### Raster Slope

```{r slope data, echo=TRUE, message=FALSE, warning=FALSE}

slope <- terrain(elev, v= "slope", neighbors=8, unit = "degrees")
sum(slope)
plot(slope, main = "Slope (unit = degrees)",col= topo.colors(50) )

#writeRaster(slope,'~/R analisi/ModLand/data/raster/slope.tiff', overwrite=TRUE)



```


#### Raster Aspect

```{r aspect, echo=TRUE, message=FALSE, warning=FALSE}

aspect <- terrain(elev, v= "aspect", unit = "degrees", neighbors=8)
sum(aspect)
plot(aspect, main = "Aspect (unit = degrees)",col= topo.colors(50) )

#writeRaster(aspect,'~/R analisi/ModLand/data/raster/aspect.tiff', overwrite=TRUE)


```



#### Geological map

```{r Geology, echo=TRUE, message=FALSE, warning=FALSE}

geo_it <- rast("~/GIS_ModLandWGS84/raster/Carta_geo_ita.tif")

plot(geo_it, main = "Geological Map", legend = FALSE,col= topo.colors(20))


```

#### Ecopedological map


```{r Geology, echo=TRUE, message=FALSE, warning=FALSE}

ecoped_it <- rast("~/GIS_ModLandWGS84/raster/Carta_ecopedo_ita.tif")

plot(ecoped_it, main = "Soils Map", legend = FALSE,col= topo.colors(20))


```


#### Global Horizontal Irradiance (GHI)

```{r GHI, echo=TRUE, message=FALSE, warning=FALSE}

GHI <- rast("~/GIS_ModLandWGS84/raster/GHI.tif")
GHI <- crop(GHI, sa, mask = TRUE)

col <- rev(heat.colors(999))
plot(GHI, main = "Global Horizontal Irradiation (GHI)", col= col)

#writeRaster(GHI,'~/R analisi/ModLand/data/raster/GHI.tiff', overwrite=TRUE)


```



